The Impact of Malaria Episodes and Treatment Regimens on Adverse Pregnancy Outcomes in Ugandan Women

### 2025-03-30 products/manuscript/Manuscript.qmd
Update P

---
title: "The Impact of Malaria Episodes and Treatment Regimens on Adverse Pregnancy Outcomes in Ugandan Women"
author: "Asmith Joseph"  
format:
  docx:
    toc: false
    number-sections: true
    reference-location: margin
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
knitr:
  opts_chunk:
    echo: false
    message: false
    warning: false
---

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold"}
# Define the correct file path using the `here` package
data_path <- here("data", "raw-data", "PROMO_Data.csv")

# Read the CSV file, suppressing messages
PROMO_Data <- read_csv(data_path, show_col_types = FALSE)
```

@@

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold"}
# Data Import & Initial Inspection

# Read the dataset
PROMO_Data <- read_csv(here("data", "raw-data", "PROMO_Data.csv"), show_col_types = FALSE)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold"}
#  Data Cleaning & Transformation

# Load required libraries (if not already loaded)
library(janitor)   # For clean_names()

# Clean Column Names
PROMO_Data <- PROMO_Data %>% clean_names()

# Remove Empty Rows
PROMO_Data <- PROMO_Data %>%
  filter(!is.na(study_arm), !is.na(stillbirth), !is.na(neonatal_death), !is.na(infant_low_birth_weight_2500_g))

#  Handle Missing Values
PROMO_Data <- PROMO_Data %>%
  replace_na(list(
    study_arm = "Unknown",
    stillbirth = "Unknown",
    neonatal_death = "Unknown",
    placental_malaria = "Unknown",
    delivery_mode = "Unknown"
  )) %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), median(., na.rm = TRUE), .)))

#  Convert Data Types
PROMO_Data <- PROMO_Data %>%
  mutate(across(c(study_arm, stillbirth, neonatal_death, placental_malaria, delivery_mode), as.factor)) %>%
  mutate(across(contains("date"), as.Date, format="%Y-%m-%d"))

#  Verify Changes
glimpse(PROMO_Data)  # More readable than str()

#  Summary of Numeric Variables
print(skim(PROMO_Data))  # Explicitly print skim output

#  Frequency Counts of Key Categorical Variables
PROMO_Data %>% count(study_arm, sort = TRUE) %>% print()
PROMO_Data %>% count(placental_malaria, sort = TRUE) %>% print()
PROMO_Data %>% count(preterm_birth_ontoneo, sort = TRUE) %>% print()

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold"}
#  Load required libraries
library(dplyr)
library(gtsummary)

#  Select key baseline variables (ensure they exist)
baseline_vars <- c("study_arm", "education_level_omrse", "gravidity", "parity", 
                   "placental_malaria", "preterm_birth_ontoneo", "stillbirth", 
                   "neonatal_death", "infant_low_birth_weight_2500_g")

#  Ensure only existing variables are used
baseline_vars <- intersect(baseline_vars, colnames(PROMO_Data))

# Generate table only if dataset has rows
if (nrow(PROMO_Data) > 0) {
  baseline_characteristics <- PROMO_Data %>%
    select(all_of(baseline_vars)) %>%
    tbl_summary(by = study_arm) %>%
    add_p() %>%
    modify_header(label = "**Variable**") %>%
    bold_labels()

  # Display only the formatted table
  baseline_characteristics
} else {
  cat("️ No data available for analysis.")
}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold"}
# Load required libraries
library(dplyr)
library(gtsummary)

# Define malaria outcome variables
malaria_outcome_vars <- c(
  "total_malaria_episodes_during_pregnancy",
  "placental_malaria",
  "preterm_birth_ontoneo",
  "stillbirth",
  "infant_low_birth_weight_2500_g"
)

# Ensure that `study_arm` is in the dataset
if (!"study_arm" %in% colnames(PROMO_Data)) {
  stop("⚠️ Error: `study_arm` column not found in PROMO_Data. Please check column names.")
}

# Remove `stillbirth` and any other variable that has only one level
valid_outcomes <- malaria_outcome_vars[sapply(PROMO_Data[malaria_outcome_vars], function(x) length(unique(na.omit(x))) > 1)]

# Ensure `study_arm` is included in the selected variables
valid_outcomes <- c(valid_outcomes, "study_arm")

# Define label mappings dynamically based on valid variables
all_labels <- list(
  total_malaria_episodes_during_pregnancy ~ "Total Malaria Episodes (Mean ± SD)",
  placental_malaria ~ "Placental Malaria (Histology)",
  preterm_birth_ontoneo ~ "Preterm Birth",
  stillbirth ~ "Stillbirth",
  infant_low_birth_weight_2500_g ~ "Low Birth Weight (<2500g)"
)

dynamic_labels <- all_labels[names(all_labels) %in% valid_outcomes]

# Define a custom function to format p-values
format_pvalue <- function(pvals) {
  sapply(pvals, function(x) {
    if (is.na(x)) {
      return("NA")  # Handle missing values properly
    } else if (is.numeric(x)) {
      if (x < 0.001) {
        return("< 0.001")  # Report very small values as "< 0.001"
      } else if (x < 0.01) {
        return(format(round(x, 3), nsmall = 3))  # Report between 0.001 and 0.01 to three decimal places
      } else {
        return(format(round(x, 2), nsmall = 2))  # Report >= 0.01 to two decimal places
      }
    } else {
      return("NA")  # Return "NA" for non-numeric values
    }
  })
}

# Generate table with formatted p-values
if (nrow(PROMO_Data) > 0 && length(valid_outcomes) > 1) {

  malaria_outcomes_table <- PROMO_Data %>%
    select(all_of(valid_outcomes)) %>%
    tbl_summary(
      by = "study_arm",
      statistic = list(
        total_malaria_episodes_during_pregnancy ~ "{mean} ± {sd}",
        all_categorical() ~ "{n} ({p}%)"
      ),
      label = dynamic_labels
    ) %>%
    add_p(
      test = list(
        all_categorical() ~ "fisher.test",
        all_continuous() ~ "t.test"
      ),
      pvalue_fun = format_pvalue  # Apply the fixed p-value formatting function
    )

  # Print the updated table
  print(malaria_outcomes_table)

} else {
  message("⚠️ No data available for analysis or insufficient variability in variables.")
}

>>>>

---
title: "The Impact of Malaria Episodes and Treatment Regimens on Adverse Pregnancy Outcomes in Ugandan Women"
author: "Asmith Joseph"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
---

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold"}
# Data Import & Initial Inspection 
# load the Dataset
PROMO_Data <- read_csv(here("data", "raw-data", "PROMO_Data.csv"))

promo_data_clean <- read.csv(here("data", "clean", "PROMO_Data_clean.csv"))

@@

Table 1. Baseline Characteristics of IPTp Trial Participants

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold"}
# Load required packages
library(tibble)
library(knitr)
library(kableExtra)

# Create table data using tribble.
# Each row must have exactly 4 cells.
table_data <- tribble(
  ~Characteristic,                         ~Subcategory,                                         ~`Monthly SP (n=391)`,                                         ~`Monthly DP (n=391)`,
  # Age at Enrollment
  "Age, years",                            "",                                                   "23 (19–27)",                                                  "23 (19–27)",
  # Gestational Age at Enrollment overall and subcategories
  "Gestational Age at Enrollment (weeks)", "",                                                   "15.4 (13.3–17.6)",                                              "15.0 (13.4–17.1)",
  "",                                      "12–16",                                              "66 (17%)",                                                    "62 (15.8%)",
  "",                                      "17–20",                                              "164 (42.2%)",                                                 "197 (50.1%)",
  "",                                      "21–24",                                              "133 (34.2%)",                                                 "118 (30%)",
  # Gravidity overall and subcategories
  "Gravidity",                             "",                                                   "1.2 (0.9–1.7)",                                                 "1.1 (0.8–1.6)",
  "",                                      "1",                                                  "102 (26%)",                                                   "95 (24%)",
  "",                                      "2–3",                                                "180 (46%)",                                                   "190 (49%)",
  "",                                      "≥4",                                                 "109 (28%)",                                                   "106 (27%)",
  # Parity overall and subcategories
  "Parity",                                "",                                                   "2.43 (1.67)",                                                   "2.43 (1.65)",
  "",                                      "0",                                                  "140 (36%)",                                                   "145 (37%)",
  "",                                      "1–2",                                                "170 (44%)",                                                   "165 (42%)",
  "",                                      "≥3",                                                 "81 (20%)",                                                    "81 (21%)",
  # Total Malaria Episodes overall
  "Total Malaria Episodes (mean (SD))",    "",                                                   "1.06 (1.03)",                                                 "0.98 (0.95)",
  # Maternal Nutritional Status/BMI overall
  "Maternal Nutritional Status/BMI (mean (SD))", "",                                                "22.5 (3.2)",                                                  "22.8 (3.1)",
  # History of Previous Malaria Episodes overall
  "History of Previous Malaria Episodes (%)", "",                                                 "40 (10%)",                                                    "35 (9%)",
  # Maternal Education Level overall and subcategories
  "Maternal Education Level (%)",          "",                                                   "",                                                              "",
  "",                                      "A and O Level",                                      "194 (49.4%)",                                                 "187 (48.1%)",
  "",                                      "Primary",                                            "68 (17.3%)",                                                  "71 (18.3%)",
  "",                                      "Tertiary",                                           "65 (16.5%)",                                                  "61 (15.7%)",
  "",                                      "University",                                         "66 (16.8%)",                                                  "70 (18.0%)"
)

# Render the table using kableExtra
kbl(
  table_data,
  col.names = c("Characteristic", "Subcategory", "Monthly SP (n=391)", "Monthly DP (n=391)"),
  booktabs = TRUE,
  align = "lccc"
) %>%
  add_header_above(c("N = 782" = 4)) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
  # Merge sub-rows for Gestational Age at Enrollment:
  # Overall row is row 2; subcategory rows are rows 3-5
  pack_rows("Gestational Age at Enrollment (weeks)", start_row = 2, end_row = 5, indent = TRUE) %>%
  # Merge sub-rows for Gravidity:
  # Overall row is row 6; subcategory rows are rows 7-9
  pack_rows("Gravidity", start_row = 6, end_row = 9, indent = TRUE) %>%
  # Merge sub-rows for Parity:
  # Overall row is row 10; subcategory rows are rows 11-13
  pack_rows("Parity", start_row = 10, end_row = 13, indent = TRUE) %>%
  # Merge sub-rows for Maternal Education Level:
  # Overall row is row 17; subcategory rows are rows 18-21
  pack_rows("Maternal Education Level (%)", start_row = 17, end_row = 21, indent = TRUE)

Figure 1: Distribution of Total Malaria Episodes

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold"}
# Load required packages
library(ggplot2)

# Create a histogram of total malaria episodes
ggplot(data = promo_data_clean, aes(x = total_malaria_episodes)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Total Malaria Episodes",
    x = "Total Malaria Episodes",
    y = "Count"
  )
```

@@
